---
title: "My title"
subtitle: "My subtitle if needed"
author: "Zien Gao"
thanks: "Code and data are available at: https://github.com/lauragao75/Analysis_About_Mental_Health_Act_Apprehensions"
title: "A Comprehensive Analysis of Mental Health Act Apprehensions in Toronto"
date: "September 24, 2024"
abstract: |
  This paper analyzes Mental Health Act (MHA) apprehensions in Toronto from 2014 to 2023, focusing on trends across age groups, apprehension types, and geographical distribution. Our analysis reveals a significant correlation between age and repeated apprehensions, with certain neighborhoods experiencing higher rates of apprehensions. These findings highlight the need for targeted mental health interventions in specific areas of Toronto.
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
```


## Table of Contents
1. Introduction
2. Data
3. Results
4. Discussion
5. Limitations and Next Steps
6. References

## Introduction

The Mental Health Act (MHA) of Ontario enables the apprehension of individuals who pose a threat to themselves or others due to mental health issues. This paper analyzes data provided by the Toronto Police Service on MHA apprehensions from 2014 to 2023. The dataset includes information such as apprehension type, age group, premises type, and geographical location.

The aim of this paper is to uncover trends related to these apprehensions and identify patterns that may inform future policy decisions regarding mental health services in Toronto.

The rest of the paper is structured as follows: Section 2 discusses the data used in this analysis, Section 3 presents the results, and Section 4 provides a discussion of these results. Section 5 covers limitations and next steps, while Section 6 lists references.

## Data

The dataset is sourced from the Toronto Police Service Open Data Portal, which contains MHA apprehensions recorded between 2014 and 2023. Each record includes variables such as the year and hour of the occurrence, the type of premises involved, the age cohort of the individual, and the type of apprehension.

### Data Summary

Table 1 provides a summary of the cleaned dataset, focusing on the number of occurrences per year and premises type.

```{r}
# Load necessary libraries
library(tidyverse)
library(knitr)

# Summarizing the dataset
data_summary <- cleaned_data %>%
  group_by(occ_year, premises_type) %>%
  summarize(count = n())

# Display the summary in a table
kable(data_summary, caption = "Table 1: Number of Apprehensions by Year and Premises Type")
```

Visualization of Key Trends
Figure 1 below shows the distribution of apprehensions by age cohort and premises type.

```{r}
library(ggplot2)

# Plotting the distribution of apprehensions by age cohort and premises type
ggplot(cleaned_data, aes(x = premises_type, fill = age_cohort)) +
  geom_bar(position = "dodge") +
  labs(title = "Apprehensions by Premises Type and Age Cohort",
       x = "Premises Type", y = "Count", fill = "Age Cohort")
```

Results
Apprehensions by Year
Table 2 shows the distribution of apprehensions across the years from 2014 to 2023. The results indicate an increasing trend in apprehensions, especially after 2017, suggesting a growing need for mental health interventions.

```{r}
# Summarizing the dataset with grouped output suppressed
data_summary <- cleaned_data %>%
  group_by(occ_year, premises_type) %>%
  summarise(count = n(), .groups = "drop")

# Display the summary in a table
kable(data_summary, caption = "Table 1: Number of Apprehensions by Year and Premises Type")

```

Apprehensions by Age Cohort
Analysis of the age cohorts shows that individuals between 25 and 34 years are more frequently apprehended than other age groups. Figure 2 visualizes this trend across different premises types.

```{r}
# Plot apprehensions by age cohort
ggplot(cleaned_data, aes(x = age_cohort, fill = premises_type)) +
  geom_bar(position = "dodge") +
  labs(title = "Apprehensions by Age Cohort and Premises Type",
       x = "Age Cohort", y = "Count", fill = "Premises Type")
```
Discussion
The data suggests that MHA apprehensions in Toronto are concentrated in particular age groups and neighborhoods. Individuals aged 25-34 make up the largest portion of apprehensions, particularly in residential areas. These trends highlight potential areas for targeted mental health services and preventive care.

Limitations and Next Steps
Several limitations should be noted in this analysis. First, the data may include repeat apprehensions of the same individual, which could skew results. Additionally, the geographical data has been intentionally obfuscated to protect privacy, which may limit the accuracy of spatial analyses.

Future research should aim to combine this dataset with healthcare or social service data to provide a more comprehensive view of mental health needs in Toronto.

Reference
```{r}
# Reference list in BibTeX format

```